---
title: "Plotting Growth Curves"
author: "UZH n UU: Who is the author here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=TRUE, echo=FALSE}
library(readr) # this package is for loading files (e.g. tsv, csv, ..)

```


# Introduction

The data that we are going to use for this exercise is published in the mBio paper [@niggli2023mbio]. 
It can be found online on: https://doi.org/10.6084/m9.figshare.23578086.v1
It describes the OD600 (growth) of the *Staphylococcus aureus* strain **JE2** in different media.

# Reading in Data and Plotting Growth Curves

```{r load data, eval=TRUE, include=TRUE, echo=TRUE}
fN <- "JE_growth_final.txt" # us this txt file that is copy pasted spread sheet from excel tab separated 
dat <- read.table(fN, header = T, sep = "\t", stringsAsFactors = TRUE)
```

```{r print data, eval=TRUE, include=FALSE, echo=TRUE}
# this print command is only shown on the html if include = TRUE
print("these are so called junks of code, you may include them to be shown in the rendered document (html here) each junk has a unique name next to it and then some additional arguments")
print(fN)
```

```{r dimensions, eval=TRUE, include=TRUE, echo=FALSE}
# 
dd <- dim(dat)
```

The dimensions of this dataset are shown above.
It is easy possible to show also content of variables in the html part.  
Note: if you end a line with 2 spaces you go to a new paragraph.  
  
Like here: `r dd`


# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```


# References {-}

