---
title: "Plotting Growth Curves"
author: "UZH n UU: Who is the author here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=TRUE, echo=FALSE}
library(readr) # this package is for loading files (e.g. tsv, csv, ..)
library(tidyverse)

```


# Introduction

The data that we are going to use for this exercise is published in the mBio paper [@niggli2023mbio]. 
It can be found online on: https://doi.org/10.6084/m9.figshare.23578086.v1
It describes the OD600 (growth) of the *Staphylococcus aureus* strain **JE2** in different media.

# Reading in Data and Plotting Growth Curves

```{r load data, eval=TRUE, include=TRUE, echo=TRUE}
fN <- "JE_growth_final.txt" # us this txt file that is copy pasted spread sheet from excel tab separated 
dat <- read.table(fN, header = T, sep = "\t", stringsAsFactors = TRUE)
```

```{r print data, eval=TRUE, include=FALSE, echo=TRUE}
# this print command is only shown on the html if include = TRUE
print("these are so called junks of code, you may include them to be shown in the rendered document (html here) each junk has a unique name next to it and then some additional arguments")
print(fN)
```

```{r dimensions, eval=TRUE, include=TRUE, echo=FALSE}
# 
dd <- dim(dat)
print(dd)
```

The dimensions of this dataset are shown above.
It is easy possible to show also content of variables in the html part.  
Note: if you end a line with 2 spaces you go to a new paragraph.  
  
Like here: `r dd`

```{r dimensions2, eval=TRUE, include=TRUE, echo=TRUE}
library(ggplot2)
ggplot(dat, aes(x = Time, y = OD600_blank_corrected, color = Growth_Medium)) + geom_point() +
  geom_smooth(method = "loess", size = 1.2) + 
  labs(title = "Growth Curves (Smoothed Regression)",
       x = "Time (hours)",
       y = "OD600 (Blank Corrected)",
       color = "Growth Medium") +
  theme_minimal()
```

```{r dimensions3, eval=TRUE, include=TRUE, echo=TRUE}
anova_result <- aov(OD600_blank_corrected ~ Growth_Medium, data = dat)
summary(anova_result)
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)
```

```{r dimensions4, eval=TRUE, include=TRUE, echo=TRUE}
qmat <- read.table("~/Downloads/r studio stuffs/omics 3/week2/challenge/qmat_1198.txt")
ggplot(qmat, )
```

# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```


# References {-}

