
---
title: "Week 2: Starting with R"
author: "Zahran Kamal (lunarlores)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=FALSE}
# install readr from CRAN
install.packages("readr")
library(readr) # make sure your R has this library loaded, if not install it (see line above)
```

# Exercises for the R-Beginner - The fancy calculator

## Exercise 2
* from https://alexd106.github.io/intro2R/exercise_2.html

### Q2: Fancy Calculator
```{r, echo=TRUE, include=TRUE }
myNumber <- 12.43

# do some calculations
# natural log
log(myNumber)
#log base 10
log10(myNumber)
#log base 2
log2(myNumber)
```
#Q3: Area determination
```{r, echo=TRUE, include=TRUE}
d <-20
area_circle <- pi*(d/2)^2
area_circle
```
#Q4: Cuberoot 
```{r, echo=TRUE, include=TRUE}
number_to_cube_root <- 14*0.51
cuberoot <- (number_to_cube_root)^(1/3)
cuberoot
```
#Q5: Vectors
```{r, echo=TRUE, include=TRUE}
#vector: https://intro2r.com/data-structures.html#scal_vecs

weight <- c(69, 62, 57, 59, 59, 64, 56, 66, 67, 66 )
weight
```
#Q6: Vector calculations
```{r, echo=TRUE, include=TRUE}
#vector: https://intro2r.com/data-structures.html#scal_vecs

weight <- c(69, 62, 57, 59, 59, 64, 56, 66, 67, 66 )
weight[2] #62
#how to work with vectors: https://intro2r.com/vectors.html

mean_weight <- mean(weight, na.rm =TRUE) #na.rm will ignore NA or missing values
median_weight <- median(weight, na.rm = TRUE)
#mode does not have a built in function in R rather it gives the type of object
type_weight <- mode(weight)
mean_weight
median_weight
type_weight

#variance of sample

sample_size <- length(weight)
sample_size

var(weight) 
#if population, var(weight)*sample_size-1/sample_size

```
#Q7: height vector
```{r, echo=TRUE, include=TRUE}
height <- c(112, 102, 83, 84, 99, 90, 77, 112, 133, 112)
summary(height) #min,max,quartile, median,mean

#positional indexes
child2_height <- height[2]
child2_height

child3_height <- height[3]
child3_height

child9_height <- height[9]
child9_height

child10_height <- height[10]
child10_height

some_child <- height[c(2,3,9,10)]
some_child

#logical indexes

shorter_child <- height[height<=99]
shorter_child
```
#Q8: BMI by vectorization
```{r, echo=TRUE, include=TRUE}
#first, unit adjustment
height_m <- height/100
BMI <- weight/height_m^2
BMI
```
#Q9: Sequences

```{r, echo=TRUE, include=TRUE}
seq1 <- seq(from =0, to = 1, by = 0.1)
seq1
```
#Q10: Sequences reversed

```{r, echo=TRUE, include=TRUE}
seq2 <- rev(seq(from =1, to = 10, by = 0.5))
seq2
```
#Q11: Sequences pro

```{r, echo=TRUE, include=TRUE}
seq3 <-rep(1:3, times=3) 
seq3

lowercase_alphabet <- c(letters)

seq4 <-rep(lowercase_alphabet[c(1,3,5,7)], each=3)
seq4

seq5 <- rep(lowercase_alphabet[c(1,3,5,7)], times=3)
seq5

seq6 <- rep(rep(1:3, each=3), times=2)
seq6

seq7 <- rep(1:5, times=rev(1:5))
seq7

seq8 <- rep(c(7,2,8,1), times=c(4,3,1,5))
seq8
```
#Q12: sort height 

```{r, echo=TRUE, include=TRUE}
height_sorted <- sort(height,decreasing = TRUE)
height_sorted
```
#Q13: childnames

```{r, echo=TRUE, include=TRUE}

child_name <- c("Alfred", "Barbara", "James", "Jane", "John", "Judy", "Louise", "Mary", "Ronald", "William")

child_name
```
#Q14: child sorting height

```{r, echo=TRUE, include=TRUE}
height_ord <- order(height)
height
height_sorted
height_ord

names_sort <- child_name[height_ord]
names_sort

```
#Q14: child sorting weight

```{r, echo=TRUE, include=TRUE}
weight_ord <- order(rev(weight))
weight
weight_ord

names_sort_weight <- child_name[weight_ord]
names_sort_weight

```
#Q16: missingdata
```{r, echo=TRUE, include=TRUE}
mydata = c(2, 4, 1, 6, 8, 5, NA, 4, 7)

mean(mydata)
mean(mydata, na.rm = TRUE)
```
#Q17: all variables
```{r, echo=TRUE, include=TRUE}

rm(seq1)
ls()
```
# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```


