---
title: "Exercise 5"
author: "Alicia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=FALSE}

library(readr)
library(tidyverse)
library(ggfortify)
```

### Q1
```{r myN, echo=TRUE, include=TRUE}
#file.choose()
dd <- read.csv("/Users/lish/Desktop/data_omics/prawnGR.CSV")
head(dd)

ggplot(dd, aes(x=diet, y=GRate)) +
    geom_boxplot(alpha = 0.7) + 
  geom_jitter(aes(color = diet), width = 0.2, alpha = 0.5) +  # Adds individual points
  theme_minimal() + 
  labs(x = "Diet Type", y = "Growth Rate", title = "Growth Rate of Prawns by Diet")

#observation nr diet type
dd %>%
  count(diet)
```

### Q2

```{r 2, echo=TRUE, include=TRUE}
#t-test
shapiro.test(dd$GRate[dd$diet == 'Natural'])
shapiro.test(dd$GRate[dd$diet == 'Artificial'])

#equal variances test
var.test(dd$GRate[dd$diet == 'Natural'], dd$GRate[dd$diet == 'Artificial'])

#both seem normally distributed and variances appear equal
#could proceed with a student's t-test
```
### Q3

```{r 3, echo=TRUE, include=TRUE}
#could proceed with a student's t-test
# H0 = no significant difference in growth rates
t.test(dd$GRate ~ dd$diet, var.equal = TRUE)
```
t-statistic = how many standard deviations the sample means are apart
df = sample sizes and assumption of equal variances

t= -1.3259:                        mean of the Artificial diet group is less than that of the Natural diet
p = 0.1901:                        H0 not rejected, no significant difference in growth rates
95% CI: (-0.613962, 0.1283495):   includes 0, supports that no signf. difference


### Q4

```{r 4, echo=TRUE, include=TRUE}
m1 <- lm(GRate ~ diet, data=dd)
summary(m1)
```

### Q5

```{r 5, echo=TRUE, include=TRUE}
anova(m1)

```
comparing t-test and anova:
- they are mathematically equivalent when comparing means of groups
- p-values (0.1901) are therefore the same
- df=1: one factor (diet) with two levels
- df=58: 60 observations minus two parameters
- F(1, 58) = 1.7579, p = 0.1901: no statistically significant difference in the mean growth rates
-> diet type did not have a measurable impact

### Q6

```{r 6, echo=TRUE, include=TRUE}
autoplot(m1)

```
assumptions of normality, linearity, and equal variance are met, minor deviations in the Q-Q plot and a few potential outliers that likely don't affect model validity


### Q7

```{r 7, echo=TRUE, include=TRUE}
#file.choose()
dd2 <- read.csv("/Users/lish/Desktop/data_omics/Gigartina.CSV")


# number replicates per diatom treatment
replics <- dd2 %>%
  count(diatom.treat)
print(replics)

# diameters across treatments
ggplot(dd2, aes(x = diatom.treat, y = diameter, fill = diatom.treat)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  labs(
    x = "Diatom Treatment",
    y = "Diameter",
    title = "Comparison of Diameters Across Diatom Treatments"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none")

```
### Q8, 9, 10

```{r 8, echo=TRUE, include=TRUE}
m2 <- lm(diameter ~ diatom.treat, data=dd2)
summary(m2)
anova(m2)

```
H0 = There is no significant difference in the mean diameter of Metacarpus

- p=1.92e-08:     at least one treatment group significantly different diam.
- Between groups (treatments): 3df
- Within groups (residuals): 36df

significant difference in the mean diameter of Metacarpus across the four treatment groups, F(3, 36) = 22.775, p < 0.001.


### Q11

```{r 11, echo=TRUE, include=TRUE}
autoplot(m2)

```
acceptable


### Q12, 13, 14

```{r 12, echo=TRUE, include=TRUE}
tukey <- TukeyHSD(aov(m2))
print(tukey)
plot(tukey, las = 1)

```
- ASGM differs from all treatments (p < 0.01).  
- Sstat differs from Sexpo (p = 0.0016).  
- No difference: Sexpo vs Sdecl, Sstat vs Sdecl (p > 0.05).  
- ASGM and Sstat show the largest differences.


### Q15

```{r 15, echo=TRUE, include=TRUE}
#file.choose()
dd3 <- read.csv("/Users/lish/Desktop/data_omics/TemoraBR.CSV")
str(dd3)
summary(dd3)
colnames(dd3)
ncol(dd3) #nr variables
nrow(dd3) #nr observations
```
dependent: beat rate
explanatory: temp

### Q16

```{r 16, echo=TRUE, include=TRUE}
dd3$Facclimatisation_temp <- as_factor(dd3$acclimitisation_temp)
str(dd3)


#coplot
coplot(beat_rate ~ temp | Facclimatisation_temp, 
       data = dd3, 
       panel = panel.smooth,
       xlab = "Temperature (°C)", 
       ylab = "Beat Rate",
       col = "blue",
       pch = 16)

#ggplot
ggplot(dd3, aes(x = temp, y = beat_rate, color = Facclimatisation_temp)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Facclimatisation_temp) +
  labs(title = "Beat Rate vs. Temperature",
       x = "Temperature (°C)",
       y = "Beat Rate",
       color = "Acclimatisation Temp")
```
rate of increase in beat rate with temp appears steeper in 5,10 than in 20:
suggests acclimatisation temp influences intercept and also rate at which beat rate changes with temperature.


### Q17, 18, 19

```{r 17, echo=TRUE, include=TRUE}
#slopes were not parallel, interaction term!
m4 <- lm(beat_rate ~ temp * Facclimatisation_temp, data = dd3)
summary(m4)
anova(m4)
```
- effect of temperature on beat_rate varies depending on the acclimatisation temperature
- acclimatisation temperature modifies the effect of temperature on beat rate
- explains why lines in facet plot not parallel
- main effects should not be interpreted in isolation

-> we could run separate models for each acclimatisation temperature

### Q20

```{r 18, echo=TRUE, include=TRUE}
autoplot(m4)
```
- no major violations regression assumptions
- slight heteroscedasticity at higher fitted values
- minor non-normality
- no strong outliers

### Q21


- Linear model with interaction term tested effects of temp and Facclimatisation_temp on beat_rate
- Significant interaction( p < 0.001) -> Effect of temp on beat rate varies by acclimatisation temp 
- Main effects significant (p < 0.001) but not interpretable alone due to interaction  
- High model fit (Adjusted R² = 0.9622)  
- Residual diagnostics: slight heteroscedasticity, no major violations  
- Beat rate increases with temp, but rate of increase depends on acclimatisation temp  
- Next step: Separate regressions for each acclimatisation group may be necessary


# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```


