---
title: "Week 2: Starting with R"
author: "UZH n UU: tested by Jonas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=FALSE}
# install readr from CRAN
# install.packages("readr")
library(readr) # make sure your R has this library loaded, if not install it (see line above)
```

# Exercises for the R-Beginner - The fancy calculator

## Exercise 2
* from https://alexd106.github.io/intro2R/exercise_2.html
```{r myN, echo=FALSE, include=TRUE}
myNumber <- 12.43

# do some calculations
# natural log
log(myNumber)
```

```{r lg10, echo=TRUE, include=TRUE}
# log10
log10(myNumber)
# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```

```{r log2, echo=TRUE, include=TRUE}
log2(myNumber)
```

```{r sqrt, echo=TRUE, include=TRUE}
sqrt(myNumber)
```

```{r exponent, echo=TRUE, include=TRUE}
exp(myNumber)
```

```{r cricle, echo=TRUE, include=TRUE}
area_circle <- pi*(10^2)
```

```{r weight, echo=TRUE, include=TRUE}
weight <- c(62, 57, 59, 59, 64, 56, 66, 67, 66)
```

```{r mean, echo=TRUE, include=TRUE}
mean(weight)
```

```{r variance, echo=TRUE, include=TRUE}
var(weight)
```

```{r sd, echo=TRUE, include=TRUE}
sd(weight)
```

```{r range, echo=TRUE, include=TRUE}
range(weight)
```

```{r lenght, echo=TRUE, include=TRUE}
length(weight)
```

```{r first five, echo=TRUE, include=TRUE}
first_five <- weight[c(1,2,3,4,5)]
```

```{r height,echo= TRUE, include=TRUE}
height <- c(112, 102, 83, 84, 99, 90, 77, 112, 133, 112)
```

```{r summary, echo=TRUE, include=TRUE}
summary(height)
```

```{r some children, echo=TRUE, include=TRUE}
some_child <- height[c(2,3,9,10)]
```

```{r shorter, echo=TRUE, include=TRUE}
shorter_child <- height[height<99]
```

```{r bmi, echo=TRUE, include=TRUE}
bmi <- weight*(height/100)
```

```{r seq1, echo=TRUE, include=TRUE}
seq1 <- seq(from=0, to=1, by=0.1)
```

```{r seq2, echo=TRUE, include=TRUE}
seq2 <- seq(from=10, to=1, by=-0.5)
```


