---
title: "Week 2: Starting with R"
author: "Zahran Kamal"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: bibliography.bib
---

```{r load libraries, include=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com/"))

# Install readr from CRAN
install.packages("readr")

# Load the library
library(readr)
```

#Q1
R script or R markdown?
#Q2
Data in "C:\Users\zasee\Documents\OmicsInOncology_CPS341nBio435_FS25_b\week2\data4exercises"

#Q3-6: opening the file here
##install
```{r, echo=TRUE, include=TRUE}
install.packages("readr")  
library(readr)
  

```
##read and display
```{r, echo=TRUE, include=TRUE}
whale_data <- read_tsv("../data4exercises/mywhaledata.tsv")
# the output tells u how many observations (rows) and variables (columns) it has, the type of columns (chr - text, dbl - numeric), and what the delimiter is

head(whale_data) #to show some columns. you can also check tail, or view for whole

str(whale_data) #if u want to see the structure

```

- These data were collected during two research cruises in the North Atlantic in May and October 2003. 
- During these two months the research vessel visited multiple stations (areas) and marine mammal observers recorded the number of whales (who doesn’t love whales!) at each of these stations. 
- The time the vessel spent at each station was also recorded along with other site specific variables such as the latitude and longitude, water depth and gradient of the seabed. 
- The researchers also recorded the ambient level of sub-surface noise with a hydrophone and categorised this variable into ‘low’, ‘medium’ or ‘high’. 
Source: https://alexd106.github.io/intro2R/exercise_3.html

#Q6-7
```{r, echo=TRUE, include=TRUE}
names(whale_data) #header values
summary(whale_data) #number_whales has 1 missing value
```
#q8a,b
```{r, echo=TRUE, include=TRUE}


whale_sub <- whale_data[1:10, 1:4]
whale_num <- whale_data[, c(1,3,4)]
whale_sub
whale_num

```
#q8c,d
```{r, echo=TRUE, include=TRUE}



whale_may <- whale_data[whale_data$month=="May",]
whale_may

whale_specific <- whale_data[11:100, 1:7]
whale_specific

```
>Remember, for some of these questions you can specify the columns you want either by position or by name. Practice both ways. Do you have a preference? If so why? 

How to do the other way?

#q9,10
```{r, echo=TRUE, include=TRUE}
whale_deeper <- whale_data[whale_data$depth>1200,]

whale_steeper <- whale_data[whale_data$gradient>200,]

whale_quieter <- whale_data[whale_data$water.noise=="low",]

whale_noisier_may <- whale_data[whale_data$water.noise=="high" & whale_data$month=="May",]

gradient_median <- median(whale_data$gradient, na.rm = TRUE)

whale_quieter_oct_steeper <- whale_data[whale_data$water.noise=="low" & whale_data$month=="October" & whale_data$gradient>gradient_median,]

between <- function(x, a, b) x >= a & x <= b


whale_faroe <- whale_data[
  whale_data$latitude >= 60.0 & whale_data$latitude <= 61.0 & 
  whale_data$longitude >= -6.0 & whale_data$longitude <= -4.0, 
]

whale_extreme <- whale_data[whale_data$water.noise!="medium",]

whale_deeper
whale_steeper
whale_quieter
whale_noisier_may
whale_quieter_oct_steeper
whale_faroe
whale_extreme
```
#Q11
```{r, echo=TRUE, include=TRUE}
whale_1500maboveavg <- whale_data[whale_data$depth>1500 & whale_data$number.whales>mean(whale_data$number.whales, na.rm = TRUE),]
mean(whale_data$number.whales, na.rm = TRUE)
whale_1500maboveavg
#mean needs to ignore NA values 
```
#Q12
```{r, echo=TRUE, include=TRUE}

may_subset <- subset(whale_data, month == "May" & time.at.station < 1000 & depth > 1000)

october_subset <- subset(whale_data, month == "October" & latitude > 61, c(1,4,5,6))

may_subset
october_subset

```
#Q13, 14
```{r, echo=TRUE, include=TRUE}
whale_depth_sort <- whale_data[order(whale_data$depth), ]
whale_depth_sort

whale_depth_noise_sort <- whale_data[order(whale_data$water.noise, -whale_data$depth), ] #minus for descending


```

#Q15 Summarise
```{r, echo=TRUE, include=TRUE}
#also use tapply()
#from guide
mean(whale_data$time.at.station)     
# mean time at station
median(whale_data$depth)             
# median depth
length(whale_data$number.whales)     
# number of observations

#aggregate
#We can also use the aggregate() function in a different way by using the formula method (as we did with xtabs()). On the left hand side of the formula (~) we specify the variable we want to apply the mean function on and to the right hand side our factors separated by a + symbol. The formula method also allows you to use the data = argument for convenience.


mean_whales_noise <- aggregate(number.whales ~ water.noise, data = whale_data, FUN = mean)
mean_whales_noise

median_whales_noise_month <- aggregate(number.whales ~ water.noise + month, data = whale_data[whale_data$month %in% c("May", "October"), ], FUN = median)
median_whales_noise_month

```
#Q16-18
tbd

# Session Information

```{r sessionInfo, echo=FALSE, include=TRUE}
pander::pander(sessionInfo())
```
